---
title: "cord clamping prj1"
output: pdf_document
members: TBD
Gruop ID: TBD
---

Data project 1, Due Feb 28.



Instructions from part1-instructions-1.pdf
https://bcourses.berkeley.edu/courses/1513233/pages/data-project


* each answer in a new page
* You must indicate on Gradescope which questions are on which pages.
* questions are pasted below


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## install pacman to help manage installation and loading of additional libraries
    
    Optional part, and it doesn't seems to play well with knit :-\
    so all code commented out for now, folx who is not into running R on your own computer don't need to run this. 
    https://www.rdocumentation.org/packages/pacman/versions/0.5.1
    pacman provides wrapper function like p_load() to install package if needed, then load the library 
    
```{r install_pacman}
    #install.packages( "pacman" )
    #library(pacman) 
    #p_load( readxl )
```








\newpage



Questions:
1. [2 marks] What is the problem your are addressing with these data? State the question you are trying
to answer and let us know what type of question this is in terms of the PPDAC framework.





\newpage

2. [2 marks] What is the target population for your project? Why was this target chosen i.e., what was
your rationale for wanting to answer this question in this specific population?




\newpage


3. [2 marks] What is the sampling frame used to collect the data you are using? Describe why you think
this sampling strategy is appropriate for your question. To what group(s) would you feel comfortable
generalizing the findings of your study and why.


\newpage

4. [2 marks] Write a brief description (1-4 sentences) of the source and contents of your dataset. Provide a
URL to the original data source if applicable. If not (e.g., the data came from your internship), provide
1-2 sentences saying where the data came from. If you completed a web form to access the data and
selected a subset, describe these steps (including any options you selected) and the date you accessed
the data.


\newpage


5. [1 mark] Write code below to import your data into R. Assign your dataset to an object.


data access: 

```{r load}

library(readxl)
library(dplyr)
cord_clamp = read_xlsx("Song_DCC_PlosOne_datadryad9.16.15.xlsx")
summary(cord_clamp)
str(cord_clamp)
set.seed(123)    # seed the pseudo random number generator, so we have consistent result between runs.
cord_clamp_sample10 = cord_clamp %>% sample_frac(0.1 )
cord_clamp_sample1  = cord_clamp %>% sample_frac(0.01)
#View(coord_clamp_sample1)
```







\newpage


6. [3 marks] Use code in R to answer the following questions:
i) What are the dimensions of the dataset?


```{r ghana_dim}
dim(cord_clamp)
```
3331 rows, 18 columns (vars)


ii) Provide a list of variable names.



They variables are listed in the docx that came with the data:
data access: 

SongDCC_PlosOne_datadryad9.16.15 ReadMe File

Studyid	
Bw – birth weight in grams
Ga – gestational age in weeks at birth
Dcc – delayed cord clamping, 1=yes, 0=no
Groups-study groups (1=DCC 30-45s, 2=DCC 60-75s)
Sex – 1= female, 2=male
Multiples = multiple pregnancy, 1=yes, 0=no
Inoutborn – 1=inborn
ans – antenatal steroids, 1=yes, 0=no
ans48 – antenatal steroids >48hours prior to delivery, 1=yes, 0=no
vag – vaginal delivery, 1=yes, 0=no
apgar1 – 1 minute apgar
apgar5 – 5 minute apgar
etippv – delivery room intubation and positive pressure ventilation, 1=yes, 0=no
drccompmeds – delivery room chest compressions and/or resuscitation medications, 1=yes, 0=no
atemp – admission temperature in Celsius
hypothermia - <36 degree Celsius on admission, 1=yes, 0=no
hct2 – hematocrit in the first 2 hours of life
hct1236 – hematocrit between 12-36 hours of life
surfactant - 1=yes, 0=no
pneumothorax - 1=yes, 0=no
intub1d – intubation within 24 hours of life, 1=yes, 0=no
anyintub – intubation any time during the NICU stay, 1=yes, 0=no
anytransfusion – any pRBC transfusion during the NICU stay, 1=yes, 0=no
pkbili – peak bilirubin, mg/dl
polycythemia – hematocrit >65, 1=yes, 0=no
died - 1=yes, 0=no	
anyivh – any intraventricular hemorrhage (grade 1-4), 1=yes, 0=no
sevivh – severe intraventricular hemorrhage (grade 3 or 4), 1=yes, 0=no
los	 - late onset sepsis, 1=yes, 0=no
anyrop – retinopathy of prematurity (stage 1-3), 1=yes, 0=no
severerop-severe retinopathy of prematurity (grade 3 or threshold disease or received anti VEGF treatment or laser treatment for ROP), 1=yes, 0=no
anynec – necrotizing enterocolitis, 1=yes, 0=no
anycld- chronic lung disease, 1=yes, 0=no
survwithoutmajormorbidity – survival without severe intraventricular hemorrhage, severe retinopathy of prematurity, late onset sepsis, necrotizing enterocolitis or chronic lung disease, 1=yes, 0=no




iii) Print the first six rows of the dataset.

```{r first_six_rows}

cord_clamp %>% head(6)

```

\newpage

7. [4 marks] Use the data to demonstrate a statistical concept from Part I of the course. Describe the
concept that you are demonstrating and interpret the findings. This should be a combination of code
and written explanation.


what do we want to present?
actually, the data should be to answer questions we want to ask with PPDAC...
so once we finalize the theme, find a data set, then define these questions?

* five numbers summary  for the bw? something else?

```{r fns}
Min    = min(       cord_clamp$bw)
Q1     = quantile(  cord_clamp$bw, probs = 0.25)
Median = median(    cord_clamp$bw)
Q3     = quantile(  cord_clamp$bw, probs = 0.75)
Max    = max(       cord_clamp$bw)


####Outliers: any data point more than 1.5*IQR away from median.
####IQR: InterQuartile Range: Q1-Median or Q3-Median

five_num_summary = cord_clamp %>% summarize( min = Min,
                                          Q1 = Q1,
                                          median = Median,
                                          Q3 = Q3,
                                          max = Max)
five_num_summary

```

* mean vs median?
*  ??


* scatter plot of bmi vs cholesterol?

```{r plot}




library(ggplot2)
ggplot(data=cord_clamp, aes(x=dcc, y=died)) + 
  geom_point(aes(col=bw))


#### maybe do a facet_wrap with ??

```


// below don't work with this dataset
```{r pressure, echo=FALSE}
plot(cord_clamp_sample10)
```


