---
title: "PHW142 Stat for Public Health, Project #3"
author:
 - Rachel Dahl
 - Sharon Drost 
 - Habiba Afrin
 - Zhigang Ma
 - Tin Ho
Gruop ID: 23
GSI: Gene Ho
Professor: Dr Mi-Suk Kang Dufour
date: "2022-04-24"

output: word_document
---

##output: pdf_document

Data project 3, Due May 2.

Title: "PHW142 Stat for Public Health, Project #3"

Subtitle: "Delayed cord clamping and preterm birth outcome" 


Team members and Student IDs (TBD):
 
Rachel Dahl 303

Sharon Drost 303

Zhigang Ma 303

Habiba Afrin 303

Tin Ho 303

Gruop ID: 23

GSI: Gene Ho

Professor: Dr Mi-Suk Kang Dufour

Top part is just copy from project 1 (part 1)  
Part 2 content starts around line 275, page 16.
Part 3 content starts around line 477

~~~~~

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Install pacman to help manage installation and loading of additional libraries
    
    https://www.rdocumentation.org/packages/pacman/versions/0.5.1
    pacman provides wrapper function like p_load() to install package if needed, then load the library 
    
```{r install_pacman}
if( !require("pacman") ) {
  install.packages( "pacman", repos = "http://cran.us.r-project.org" )
}
library(pacman) 
p_load( readxl )
p_load( gridExtra )
p_load( broom )
```

\newpage



Questions:
1. [2 marks] What is the problem your are addressing with these data? State the question you are trying
to answer and let us know what type of question this is in terms of the PPDAC framework.



PPDAC stands for Problem, Plan, Data Analysis, and Conclusion. 
Problem: What is the optimal timing of delayed cord clamping (DCC) in order to offer very preterm infants to maximize placental transfusion without interfering with stabilization and resuscitation.
Outcomes that were observed in the study we selected for data included: neonatal death, intraventricular hemorrhage, hematocrit level, delivery room intubation, hypothermia, need for surfactant therapy, intubation in first 24 hrs, any intubation during hospitalization, development of chronic lung disease, late onset sepsis, any red blood cell transfusion during hospitalization, necrotizing enterocolitis, and severe retinopathy of prematurity. The hypothesis was that prolonged (60-75 sec) delayed cord clamping would offer improved outcomes compared to shorter (30-45 sec) clamping without complications of desaturation or stabilization.

This was a causative/etiological question since we wanted to know whether changing the exposure/intervention (i.e. longer cord clamping) would result in changed outcomes. 




\newpage

2. [2 marks] What is the target population for your project? Why was this target chosen i.e., what was
your rationale for wanting to answer this question in this specific population?


The target population is very preterm infants (i.e. infants born at <32 weeks gestation) in the United States spontaneously breathing at birth. Very premature infants require stabilization and are at a high risk for requiring resuscitation at birth. This target population was chosen because the location where the data was collected serves a high-risk obstetric population, where they treat a large number of very preterm infants at the Santa Clara Valley Medical Center with a regional level III Neonatal Intensive Care Unit (NICU). 
More broadly, our Prob/Stat team chose this population because improving preterm health care is a relevant and current public health problem particularly in underserved populations. 



\newpage


3. [2 marks] What is the sampling frame used to collect the data you are using? Describe why you think
this sampling strategy is appropriate for your question. To what group(s) would you feel comfortable
generalizing the findings of your study and why.


The sampling frame included very preterm infants born between January 2008 and April 2014 at the Santa Clara Valley Medical Center. The sample included a total n of 353 infants, 187 in group 30-45 seconds DCC and 166 in group 60-75 seconds DCC. 30-45 seconds group was done from 2008-2011 and 60-75 seconds group was done from 2011-2014. This sampling strategy was appropriate because the sample was large enough to meet pre-calculated
 power requirements and included very preterm infants which is the population that we wanted to study. Additionally, there were no significant differences in baseline demographic characteristics between the exposed group (60-75 seconds DCC) and unexposed group (30-45 seconds group), allowing improved comparison between outcomes with respect to interventions. The findings of this study would likely generalize to other very preterm infants spontaneously breathing at birth in a hospital setting with a level III NICU in the United States, but more studies are warranted to confirm these results. It is unclear whether these findings would generalize to outside the hospital setting, e.g. home births, given there could be confounding variables from other interventions in the hospital setting that were not analyzed.



\newpage

4. [2 marks] Write a brief description (1-4 sentences) of the source and contents of your dataset. Provide a
URL to the original data source if applicable. If not (e.g., the data came from your internship), provide
1-2 sentences saying where the data came from. If you completed a web form to access the data and
selected a subset, describe these steps (including any options you selected) and the date you accessed
the data.

Data is available through paper citation  https://www-ncbi-nlm-nih-gov.proxy.lib.uiowa.edu/pmc/articles/PMC4577121/pdf/pone.0138829.pdf 
or directly through this website https://datadryad.org/stash/dataset/doi:10.5061/dryad.4q3d3

The data came from a prospective clinical study performed in the Santa Clara Valley Medical Center, as described above, that was published in PLOS One. The hospital had developed a standardized delivery room management “bundle” or protocol in 2007 prior to the start of this study. Very preterm infants spontaneously breathing at birth were managed using this bundle with the only difference in this study being timing of delayed cord clamping, 30-45 seconds from Jan 2008-March 2011, or 60-75 seconds from March 2011-April 2014; in that way, the infants were randomized. The dataset falls into three categories: demographics, cord clamping (30-45 seconds group vs 60-75 seconds group) and neonatal outcomes data. There were no significant differences between the two infant groups with respect to gestational age, body weight at birth, receiving antenatal steroids, birth by Caesarian section, or percentage of male sex. Outcomes (as described above) were identified for each patient during their hospitalization.



\newpage


5. [1 mark] Write code below to import your data into R. Assign your dataset to an object.


Loading data: 

```{r load}

library(readxl)
library(dplyr)
song = read_xlsx("Song_DCC_PlosOne_datadryad9.16.15.xlsx")
#View(song)
```


Optional part, poking around the data to get the lay of the land

```{r feeling the data}
#summary(song)
str(song)
set.seed(123)    # seed the pseudo random number generator, so we have consistent result between runs.
```

\newpage


6. [3 marks] Use code in R to answer the following questions:
i) What are the dimensions of the dataset?


```{r data_dim}
dim(song)
```

Answer provided by R: 353 rows, 36 columns.


ii) Provide a list of variable names.


The variables are listed in the docx that came with the data:
data access: 

SongDCC_PlosOne_datadryad9.16.15 ReadMe File

Studyid	
Bw – birth weight in grams
Ga – gestational age in weeks at birth
Dcc – delayed cord clamping, 1=yes, 0=no
Groups-study groups (1=DCC 30-45s, 2=DCC 60-75s)
Sex – 1= female, 2=male
Multiples = multiple pregnancy, 1=yes, 0=no
Inoutborn – 1=inborn
ans – antenatal steroids, 1=yes, 0=no
ans48 – antenatal steroids >48hours prior to delivery, 1=yes, 0=no
vag – vaginal delivery, 1=yes, 0=no
apgar1 – 1 minute apgar
apgar5 – 5 minute apgar
etippv – delivery room intubation and positive pressure ventilation, 1=yes, 0=no
drccompmeds – delivery room chest compressions and/or resuscitation medications, 1=yes, 0=no
atemp – admission temperature in Celsius
hypothermia - <36 degree Celsius on admission, 1=yes, 0=no
hct2 – hematocrit in the first 2 hours of life
hct1236 – hematocrit between 12-36 hours of life
surfactant - 1=yes, 0=no
pneumothorax - 1=yes, 0=no
intub1d – intubation within 24 hours of life, 1=yes, 0=no
anyintub – intubation any time during the NICU stay, 1=yes, 0=no
anytransfusion – any pRBC transfusion during the NICU stay, 1=yes, 0=no
pkbili – peak bilirubin, mg/dl
polycythemia – hematocrit >65, 1=yes, 0=no
died - 1=yes, 0=no	
anyivh – any intraventricular hemorrhage (grade 1-4), 1=yes, 0=no
sevivh – severe intraventricular hemorrhage (grade 3 or 4), 1=yes, 0=no
los	 - late onset sepsis, 1=yes, 0=no
anyrop – retinopathy of prematurity (stage 1-3), 1=yes, 0=no
severerop-severe retinopathy of prematurity (grade 3 or threshold disease or received anti VEGF treatment or laser treatment for ROP), 1=yes, 0=no
anynec – necrotizing enterocolitis, 1=yes, 0=no
anycld- chronic lung disease, 1=yes, 0=no
survwithoutmajormorbidity – survival without severe intraventricular hemorrhage, severe retinopathy of prematurity, late onset sepsis, necrotizing enterocolitis or chronic lung disease, 1=yes, 0=no


```{r var_list}
colnames(song)
```

iii) Print the first six rows of the dataset.

```{r first_six_rows}
head(song,6)
```

\newpage

7. [4 marks] Use the data to demonstrate a statistical concept from Part I of the course. Describe the
concept that you are demonstrating and interpret the findings. This should be a combination of code
and written explanation.

We hypothesized that birth weight and gestational age would have a direct positive correlation. Both of these are continuous variables, as such, we made scatter plots and performed linear regressions to identify if a linear model applied to this data. We also converted the variables to a log scale to improve the visualization of the relationship.  What we found is that converting  birth weight to a log scale yielded a slightly better R-squared correlation score than leaving it in linear scale.  This is not a huge improvement, since R^2 was still only at about 0.69, but the data points in the XY scatter plots does seem to have a more uniform spread from the lm() abline in this case.

For future parts of the project, we will evaluate other relationships of the data and further investigate birth weight, gestational period, delayed cord clamping and preterm outcome. 


```{r plot}

library(ggplot2)
library(tidyverse)
library(broom)
library(gridExtra)  # provides the grid.arrange() fn to display multiple plots on same graph

song_lm         = lm(     bw  ~     ga,  song )
song_lm_logbw   = lm( log(bw) ~     ga,  song )
song_lm_logga   = lm(     bw  ~ log(ga), song )

icept = song_lm$coefficients[[1]]
slp   = song_lm$coefficients[[2]]
icept_logbw   = song_lm_logbw$coefficients[[1]]
slp_logbw     = song_lm_logbw$coefficients[[2]]
icept_logga   = song_lm_logga$coefficients[[1]]
slp_logga     = song_lm_logga$coefficients[[2]]

rsq         <- glance( song_lm         ) %>% pull(r.squared) %>% round(4)
rsq_logbw   <- glance( song_lm_logbw   ) %>% pull(r.squared) %>% round(4)
rsq_logga   <- glance( song_lm_logga   ) %>% pull(r.squared) %>% round(4)

xy = ggplot( data=song, aes(x=ga, y=bw)) +  geom_point() +
  geom_abline( intercept=icept,     slope=slp,     col="blue",  lty=2 ) + 
  labs( title=paste("R²=", rsq ), x="Gestation (week)", y="Birth Weight (g)")

xybw   = ggplot( data=song, aes(x=ga, y=log(bw))) +  geom_point() +
  geom_abline( intercept=icept_logbw, slope=slp_logbw, col="cyan", lty=1 ) +
  labs( title=paste("R²=", rsq_logbw ), x="Gestation (week)", y="Log of Birth Weight (g)")

xyga   = ggplot( data=song, aes(x=log(ga), y=bw)) +  geom_point() +
  geom_abline( intercept=icept_logga, slope=slp_logga, col="green", lty=1 ) +
  labs( title=paste("R²=", rsq_logga ), x="Log of Gestation (week)", y="Birth Weight (g)")

grid.arrange(xy,xybw,xyga, nrow=2, top="Gestational Age vs Birth Weight")

```

\newpage



# Part 2 #######################################################################



1. [1 mark] Include your work for Part I.


_see above :)_


\newpage

2. [2 marks] Describe a quantity you will estimate as an outcome in your problem using probability notation. Are you planning to calculate marginal probabilities? Conditional probabilities?


We would like to determine the probability that the patient died given s/he was exposed to either of the two different timings of Delayed Cord Clamp procedure (DCC) (Group 1 vs Group 2).  We first create a 2x2 contingency table as:


2x2                   |  Death=Yes | Death=No | Margin        |
----------------------|------------|----------|---------------|
DCC Group=1 (30-45s)  | A:    9    | B:   178 | A+B:      187 |
DCC Group=2 (60-75s)  | C:    8    | D:   158 | C+D:      166 |
Margin                | A+C: 17    | B+D: 336 | A+B+C+D:  353 |



This is a conditional probability based on exposure to shorter or longer delayed cord clamping. The deaths per groups are shown in the Table above. The probability that a patient died conditionally based on exposure type is :

P( Death=Yes | DccGroup=1 ) = a / (a+b) = 9/187 = 0.04812834 or 4.81%

P( Death=Yes | DccGroup=2 ) = c / (c+d) = 8/166 = 0.04819277 or 4.82%

Overall, we observe that the conditional probability is approximately the same in either group.


The R code to query the data to create the 2x2 contingency table is below.

```{r constructing 2x2}

# b, a:
song      %>% filter( groups == 1 ) %>% group_by( died  )  %>% tally()

# d, c:
song      %>% filter( groups == 2 ) %>% group_by( died  )  %>% tally()

# a+b, c+d:
song      %>% group_by( groups  ) %>% tally()

#  b+d, a+c: 
song      %>% group_by( died    ) %>% tally()

# a+b+c+d
song %>% tally()

```

\newpage

3. [3 marks] Describe the type of theoretical distribution that is relevant for your data.

• What type of variable(s) are you investigating (continuous, categorical, ordinal, etc)?

I. A variable is any characteristic of an individual. Categorical variables have grouping levels and Quantitative variables are numeric variables. This data set have mainly Quantitative variables such as gestational age (Ga), birth weight (Bw), hematocrit (hct2), etc.  Furthermore, Ga is a Discrete Quantitative variable meaning this variable can be counted. Bw and hct are Continuous Quantitative variables that can be measured precisely. 

\newpage

• What theoretical distribution that we have talked about would potentially be appropriate to use with these data (Normal, Binomial, Poisson. . . )

II. Data can be described using different distributions such as Normal, Binomial, and Poisson. Normal distribution is a Continuous Distribution. Binomial and Poisson are Discrete Probability Distributions. 
For our data set, Normal Distribution would potentially be appropriate to use. We will illustrate later in document.


\newpage

• Why is this an appropriate model for the data you are studying?

III. Normal Distribution is appropriate for our analysis of hematocrit levels (hct2), which is a continuous data type. 
We have mean (µ) and standard deviation (σ) for this data set, thus describing the main statistics for a Normal distribution. To further check whether our hct2 data follows a Normal distribution, we first plotted the data as a histogram.  Visually, the plot looks approximately like a symmetric bell curve, fairly well centered in the middle for mean µ and not skewed.  We acknowledge that there are a few outliers. These may be due to measurement error because the corresponding Hematocrit values would be physiologically improbable. 


```{r NormDist-for-group-consideration}

ggplot(data=song, aes(x=hct2)) +  geom_histogram( bins=30 ) 

```

Lastly, we create a QQ plot of the hct2 data, and observes that most points track closely to the main diagonal, with a few points on the extremes that would corresponds to the outliers.  Given that the bulk of the data "fits", we now have further evidence that the data follows a Normal distribution.
```{r qq plot}

ggplot(data=song, aes(sample=hct2)) + stat_qq() + stat_qq_line()  

```

\newpage

[previous newpage did not take, hopefully didn't introduce a blank page now with a 2nd newpage]

\newpage

4. [4 marks] Use the data you have to demonstrate a statistical concept from Part II of the course. Describe the concept that you are demonstrating and interpret the findings. This may include code in R, a visual of some kind and text interpretation.


We want to determine if using longer cord clamping time (for those in group 2) would be associated with a significantly higher mean hematocrit level than the mean of both delayed cord clamping groups combined. Hematocrit is defined as the percentage by volume of packed red blood cells in a given sample of blood after centrifugation.

Statistically, our Null and Alternative Hypotheses are:

H_o:   mean hct2 of group 2 = 48.25    

H_alt: mean hct2 of group 2 > 48.25 
 
Clinically, this is what our hypothesis means:
 
H_o:   mean hct2 of group 2 (longer delayed cord clamping) = mean hct2 of entire group combined (both shorter and longer DCC) = 48.25    

H_alt: mean hct2 of group 2 (longer delayed cord clamping) > 48.25 

This was a one-sided hypothesis test, and we define alpha = 0.05. 

Next we calculate the z-statistic to determine whether the mean of the longer delayed cord clamping group was significantly larger than the mean from the combined group mean.  For this, we used R to identify the means and standard deviation from each group (combined and group 2 alone).




```{r mean_vs_median}

library(dplyr)

cord_clamp_All = song 
Min    = min(       cord_clamp_All$hct2,               na.rm=T )
Q1     = quantile(  cord_clamp_All$hct2, probs = 0.25, na.rm=T )
Median = median(    cord_clamp_All$hct2,               na.rm=T )
Mean   = mean(      cord_clamp_All$hct2,               na.rm=T )
Q3     = quantile(  cord_clamp_All$hct2, probs = 0.75, na.rm=T )
Max    = max(       cord_clamp_All$hct2,               na.rm=T )
SD     = sd(        cord_clamp_All$hct2,               na.rm=T )

seven_num_summary_All =  c( min    = Min,
                            Q1     = Q1,
                            median = Median,
                            mean   = Mean,
                            sd     = SD,
                            Q3     = Q3,
                            max    = Max)


cord_clamp_G2 = song %>% filter(groups==2)
Min    = min(       cord_clamp_G2$hct2,               na.rm=T )
Q1     = quantile(  cord_clamp_G2$hct2, probs = 0.25, na.rm=T )
Median = median(    cord_clamp_G2$hct2,               na.rm=T )
Mean   = mean(      cord_clamp_G2$hct2,               na.rm=T )
Q3     = quantile(  cord_clamp_G2$hct2, probs = 0.75, na.rm=T )
Max    = max(       cord_clamp_G2$hct2,               na.rm=T )
SD     = sd(        cord_clamp_All$hct2,              na.rm=T )


seven_num_summary_G2  =  c( min    = Min,
                            Q1     = Q1,
                            median = Median,
                            mean   = Mean,
                            sd     = SD,
                            Q3     = Q3,
                            max    = Max)


round( seven_num_summary_All, 2 )
round( seven_num_summary_G2,  2 ) 

```

From the above, we observe that:

x-bar : 49.19

mu    : 48.25

Sigma : sd = 7.03

n     : 151 (group 2 count - missing data from group 2)

There were 166 patients within the longer delayed group. Fifteen hematocrit values were missing among this group, so our n = 166 - 15 = 151. 

To calculate our z-statistic, we used this equation:

z = ( (x-bar - mu) / (sigma/sqrt(n) ) 

z = (49.19-48.25)/(7.03/sqrt(151))= 1.643089

And then used the pnorm function in R to calculate the p-value:

```{r calculate p-value}
pnorm(1.643089, mean=0, sd=1, lower.tail=F) 

```

Interpretation:

Our p-value of 0.05018226 does not quite reach our definition of significance because it does not fulfill being less than 0.05. However, we think that it is borderline significant with an alpha of 0.05, meaning that we think there likely is a difference between the shorter and longer delayed cord clamping groups. Another future study with a larger sample size might be warranted to further definitively determine whether there truly is a significant benefit with longer periods of delayed cord clamping. 

\newpage

# =================== PART 3 =========================


1. [2 marks] Identify a statistical test to apply to your data (must be a concept we covered in part III of
the course). In plain language, write the question you are trying to answer.

We used ANOVA. We wanted to determine whether Apgar5 scores within categories that we defined for this project as Low (0-6), Medium (7-8) or High (9-10) ranges are associated with measured hematocrit levels at 12-36 hours after birth.

The APGAR test is a quick scoring system performed on observation of a newborn at 1 and 5 minutes after birth. The 1-minute score determines how well the baby tolerated the birthing process. The 5-minute score tells the health care provider how well the baby is doing independently outside the mother's womb. In rare cases, if the newborn is considered unstable, the test will be done 10 minutes after birth.

Hematocrit (Hct) is defined as the volume of red blood cells compared to the volume of whole blood after centrifugation. Red blood cells carry oxygen throughout your body. Having too few or too many red blood cells can be a sign of certain diseases. The Hct level would give us insight on the health of the newborn at 5 minutes after birth and is used to prognosticate the morbidity and mortality of the newborn in the first week of life.

Since Apgar5 is a categorical variable with 3 groups, and hematocrit is a continuous variable, ANOVA is a useful test for this dataset.


```{r Part 3 Q1 }
# the pacman library can install libraries (if not installed) before loading it
# but pacman can't quite install itself, so next two lines may not be usable everywhere:
#library(pacman)
#p_load(forcats)

library(forcats)

# Here we start prepping the data for our data analysis
# the data was contained by the variable "song_ext" prepared from previous assignment.
# Here we drop records that has missing values (a fact of life according to data scientists)
# apgar5 variable contain discrete scores from 0 to 10, 
# we remap them to categories Low, Med, High, stored in a column called apgar5Cat 
# We end by manually sorting the categories per human mental mind of L,M,H 
# instead of alphabetically using  fct_relevel().

song_ext = na.omit(song) %>%  
  mutate(  apgar5Cat  =   case_when(       apgar5==0  ~ "Low" , 
                                           apgar5==1  ~ "Low" , 
                                           apgar5==2  ~ "Low" , 
                                           apgar5==3  ~ "Low" , 
                                           apgar5==4  ~ "Low" , 
                                           apgar5==5  ~ "Low" , 
                                           apgar5==6  ~ "Low" , 
                                           apgar5==7  ~ "Med" , 
                                           apgar5==8  ~ "Med" , 
                                           apgar5==9  ~ "High" ,  
                                           apgar5==10 ~ "High" 
                                           ) ) %>%
  mutate( apgar5Cat = fct_relevel( apgar5Cat, "Low", "Med", "High" ))

```

\newpage

2. [2 marks] What assumptions are required by the method you chose in 5.1? Show how you assessed
whether these assumptions are met by your dataset. 

For Anova we are required to make the following assumptions:

* All samples are SRSs and independent of each other. We assume we can recover the probability of the examples that’s representative from each target group. The groups were designed in such a way that they are independent of each other.

* Each of the sample groups are normally distributed: we have QQ plots for each group, and the data points track the main diagonal reasonably well, without huge outliers.  Each group has enough sample points with n >= 30.  


```{r Part 2 Q2 check for normality}

# Here we use ggplot stat_qq to produce QQ plots


# But first gather the sample count for each categories, 
# so that we can display this data as subtitle of the plot
apgarCounts = song_ext %>% count( apgar5Cat )


# facet_wrap create 3 panels for the 3 categories of Apgar5 data of Low,Med,High
# QQ plot from Week 6 Lect 15 slide 9
qq_plot = ggplot(data=song_ext, aes(sample=hct1236)) + stat_qq() + stat_qq_line() + facet_wrap( ~ apgar5Cat) +
  labs( title="QQ plot for Hematocrit at 12-36 hours by APGAR5 categories", 
        subtitle=paste("Sample size for Low,Med,High:", 
                       apgarCounts$n[1], ", ", 
                       apgarCounts$n[2], ", ", 
                       apgarCounts$n[3]         ), 
        x="z score", 
        y="sample")

qq_plot

```


* All the populations have the same standard deviation σ.   While we do not know what is the σ of the population, all patients were drawn at random and we are not aware of any systematic segregation that would make different groups have different σ.  We are also following the rule of thumb to determine the ratio of largest vs smallest standard deviation, and found the ratio to be less than 2, thus giving us confidence that this condition is met.
SDmax/SDmin = 10.197679/6.798941= 1.4998


```{r Part 3 Q2 }

# use group_be and summarize to determine group sample size and their std dev
SD_by_Apgar_Grp = song_ext %>% group_by(apgar5Cat) %>% 
  summarize( SD = sd(hct1236),
             n  = length(hct1236))
SD_by_Apgar_Grp

# determine the ratio of the largest std dev vs the smallest
sd.ratio = SD_by_Apgar_Grp$SD[1] / SD_by_Apgar_Grp$SD[3]
sd.ratio


```


Overall, we have strong evidence to suggest that our data meets the conditions needed by the ANOVA tests and our result would be statistically rigorous.  

\newpage

3. [2 marks] Explain why this test is appropriate for the data you have and the question you are trying
to answer. Use at least one visualization technique and include both the output and the R code that
generated it.

We have 3 categories for our Apgar5 data: Low, Medium, High.  Each group has a random sample of patients with different hematocrit levels (measured at the 12-36 hours), which is a continuous variable.  We would like to see if the hct level differs between these groups.  ANOVA would allow us to evaluate multi-categories data.  It has several assumptions, mostly centered around Normality and large enough samples, which we have met.  

Let’s start by visualizing the spread of our data points:

  
```{r Part 3 Q3 }

# We will use ggplot boxplot() to show the spread of the 3 groups of data
# boxplot, with 
# jitter so points are shown with random offset to make overlapping pts visible
# subtitle data variable apgarCounts created two blocks back
plot3cats = ggplot( data=song_ext, aes(y=hct1236,x=apgar5Cat)) +  
  geom_boxplot() + geom_jitter(width=0.1) +
  labs( title="Hematocrit at 12-36 hours across APGAR5 categories", 
        subtitle=paste("Sample size for Low,Med,High:", 
                       apgarCounts$n[1], ", ", 
                       apgarCounts$n[2], ", ", 
                       apgarCounts$n[3]         ), 
        x="5 minutes Apgar score", 
        y="12-36 hours Hematocrit level")


plot3cats
```
We can see that the spread of the data among all 3 groups are about the same, which is expected since we have already found that SDmax/SDmin < 2.  The IQR of the 3 groups overlap, and visually the mean value differs slightly between them.  Whether the differences are statistically significant will be tackled next.

\newpage


4. [2 marks] Clearly state the null and alternative hypotheses for this test.

Null Hypothesis: There is no difference of mean value of Hematocrit measured at 12-36 hours (Hct1236) among the 3 levels of 5 minutes Apgar score categories (apgar5cat) of low(u1), medium(u2) and high(u3). 
H0: u1=u2=u3
The alternative hypothesis is that at least one mean value of Hct1236 is different from other groups.
Ha: u1≠u2≠u3, or u1≠u2, or u1≠ u3, or u2≠u3 

\newpage


5. [2 marks] Include the R code you used to generate your results. Annotate your code to help us follow
your reasoning.  

With the preparation code up to this point, we can now calculate the ANOVA F-score for our data:



```{r anova}

# we use aov() function to perform ANOVA test.  
# data has been prepared and stored in the song_ext data frame.  
# result of aov() is piped thru tidy for easier readability
aov( hct1236 ~ apgar5Cat, data=song_ext ) %>% tidy()
``` 

Here we see that the F-statistics is 9.59, with a very low p.value of 0.0001, which is much smaller than a typical alpha threshold of 0.05.  Thus, we have evidence to support rejecting the null.  


\newpage

6. [4 marks] Present your results in a clear summary. This should include both a text summary and a
table or figure with appropriate labeling. 

After finding there is support for the  alternative hypothesis, we proceed to do pairwise comparisons of the various apgar5 groups.  We will use the TukeyHSD() function in R to help automate this work:



```{r turkey}
TukeyHSD( aov(hct1236 ~ apgar5Cat, data=song_ext))

```
The results from Tukey HSD test performed with a 95% confidence level displayed significant differences between the medium (n=103) and low (n=36) APGAR groups and between the high (n=77) and low APGAR groups, with p<0.05 for both comparisons. There was a trend that did not quite reach significance for difference between high and medium APGAR groups, with adjusted p=0.06.


\newpage 

7. [4 marks] Interpret your findings. Include a statement about the evidence, your conclusions, and the
generalizability of your findings.  

Based on our statistical findings, we reject the null hypothesis and support the alternate hypothesis that higher APGAR scores (APGAR 5 min) are associated with higher hematocrit levels in newborns at 12-26 hours after birth. Higher hematocrit levels are associated with more favorable outcomes with regard to morbidity and mortality in this population of very preterm infants (i.e.infants born at <32 weeks gestation) in the United States spontaneously breathing at birth (Fogarty, et al., 2018). Very premature infants require stabilization and are at a high risk for requiring resuscitation at birth. This target population was chosen because the location where the data was collected serves a high-risk obstetric population, where they treat a large number of very preterm infants at the Santa Clara Valley Medical Center with a regional level III Neonatal Intensive Care Unit (NICU). More broadly, our small group chose this population because the results would generalize to similar populations. Improving preterm health care is a relevant and current public health problem particularly in underserved populations.


# References

Fogarty, M., Osborn, D. A., Askie, L., Seidler, A. L., Hunter, K., Lui, K., Simes, J., & Tarnow-Mordi, W. (2018). Delayed vs early umbilical cord clamping for preterm infants: a systematic review and meta-analysis. American journal of obstetrics and gynecology, 218(1), 1–18. https://doi-org.proxy.lib.uiowa.edu/10.1016/j.ajog.2017.10.231

Song, Dongli; Jegatheesan, Priya; DeSandre, Glenn; Govindaswami, Balaji (2016), Data from: Duration of cord clamping and neonatal outcomes in very preterm infants, Dryad, Dataset, https://doi.org/10.5061/dryad.4q3d3

